Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-08-11 text/sections/00-abstract.tex
adjusted writing

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while states have limited how much they fund higher education.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that state funding cuts affect composition of faculty for the next 3 years, showing that funding cuts have lingering effects.
Analysis of all the professors at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

>>>>

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one mechanism for worsening student outcomes at US public universities over this same time period.

### 2023-08-11 text/sections/01-introduction.tex
adjusted writing

I use data from the \cite{ipeds} to study the impact of changes in state funding for US public universities 1990--2017.
I combine these data with panel data for every public university faculty member in the state of Illinois for 2010--2021 \citep{ibhed}, to measure how faculty are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), do they change the composition of faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty members themselves?

>>>>

I use data from the \cite{ipeds} to study the impact of changes in state funding for US public universities 1990--2017.
I combine these data with panel data for every faculty member at Illinois public universities for 2010--2021 \citep{ibhed}, to measure how incumbent faculty (i.e., those already employed at the university) are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), do they change the composition of faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty members themselves?
